---
title: "Predicting Breast Cancer Metastasis Using Clinicopathological Data and Machine Learning Technologies"
author: Yi-Ju Tseng, Chuang-En Huang, Po-Yin Lai, Yu-Chen Sun, Chiao-Ni Wen, Hsin-Yao Wang, and Jang-Jih Lu
institute: "Chang Gung University"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```
## Set environment
### Load all librarys
```{r library, warning=F,message=F}
library(data.table)
library(ggplot2)
library(lubridate)
library(dplyr)
library(tidyr)
library(ROCR)
library(caret)
library(tableone)
```
### Load all functions
```{r warning=F,message=F}
# The ModelFunction.R can be download from GitHub
source('ModelFunction.R')
```

## Data load and pre-process
### Load breast cancer data for model development

```{r data.load, warning=F}
TimeIndepData<-readRDS('TimeIndepData.rds')
str(select(TimeIndepData,-ID))
TimeVariedData<-readRDS('TimeVariedData.rds')
str(select(TimeVariedData,-ID))
```

### T stage code conversion

```{r code.conversion,warning=F}
TimeIndepData[pTi %in% c('4a','4b','4c','4d'),pTi:='8']
TimeIndepData[pTi=='3',pTi:='7']
TimeIndepData[pTi=='2',pTi:='6']
TimeIndepData[pTi=='1c',pTi:='5']
TimeIndepData[pTi=='1b',pTi:='4']
TimeIndepData[pTi=='1a',pTi:='3']
TimeIndepData[pTi=='1micro',pTi:='2']
TimeIndepData[pTi=='is',pTi:='1']
TimeIndepData[pTi=='x',pTi:=NA]
```

### Process categorical variables
```{r factor}
cols <- c("pTi", "pNi", "pMi", "Tissue.ER", "Tissue.PR", "Tissue.HER2")
TimeIndepData[, c(cols) := lapply(.SD, as.factor), .SDcols= cols]
```

## Patient characteristics
```{r tableone,warning=F}
LastData <- TimeVariedData[order(ID,LabName,-RCVDAT)][,.SD[c(1)],by=.(ID,LabName)]
LabWide <- spread(LastData[,.(ID,LabName,LabValue)],key=LabName,value=LabValue)
IndexDateGap<-LastData[,.(ID,Index_duration)][order(ID,Index_duration)][,.SD[c(1)],by=.(ID)]

LastRecordAfterOP <- inner_join(TimeIndepData,LabWide, by = "ID")%>%
  inner_join(., IndexDateGap, by='ID') 

vars <- c("OPAge","pTi","pNi","pMi","Tissue.ER","Tissue.PR","Tissue.HER2","CA153","CEA","HER2","Index_duration") 
cols <- c("pTi", "pNi", "pMi", "Tissue.ER", "Tissue.PR", "Tissue.HER2")
ALL_tableOne <- CreateTableOne(vars = vars, strata = c("IsRe"),factorVars = cols, data = LastRecordAfterOP) 
```
```{r warning=F,message=F,echo=T,results='hide'}
ALL_tableOneDF<-data.table(as.matrix(print(ALL_tableOne,nonnormal= "Index_duration")),keep.rownames = T)
```
```{r warning=F,message=F}
knitr::kable(ALL_tableOneDF)
```

## Build the model to predict breast cancer metastasis at least 3 months in advance

### Set parameters
```{r}
seed<-3231
n_times<-50
n_folds<-3
trc<-trainControl(method = "cv", number = n_folds, 
                  classProbs=TRUE, summaryFunction = twoClassSummary)
```

### Get completed data 3 months before the index date
```{r data3months,warning=F,message=F}
Over90dData <- TimeVariedData %>% filter(Index_duration>90) %>% 
  arrange(ID,LabName,Index_duration) %>% group_by(ID,LabName) %>% slice(1)
LabWide90d <- spread(Over90dData %>% select(ID,LabName,LabValue),key=LabName,value=LabValue)
LastRecordAfterOP_90<-inner_join(TimeIndepData,LabWide90d, by = "ID")
LastRecordAfterOP_90_rmna <- LastRecordAfterOP_90 %>% filter(complete.cases(LastRecordAfterOP_90))
LastRecordAfterOP_90_rmna_IsRe<-LastRecordAfterOP_90_rmna %>% filter(IsRe=="Recurrence")
LastRecordAfterOP_90_rmna_NotRe<-LastRecordAfterOP_90_rmna%>% filter(IsRe=="Non.Recurrence")
nrow(LastRecordAfterOP_90_rmna_IsRe)
nrow(LastRecordAfterOP_90_rmna_NotRe)
```

### 3 evaluation folds
```{r sampling,warning=F,message=F}
datalist<-generate3folds(LastRecordAfterOP_90_rmna_IsRe,LastRecordAfterOP_90_rmna_NotRe,seed)
training_90<-datalist[[1]]
test_90<-datalist[[2]]
```

### Model tuning and evaluation - logistic regression
```{r warning=F,message=F}
glm_perf_90<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    glm_perf_90_tmp<-glm_tune_eval(training_90,test_90,i,k,seed)
    glm_perf_90<-rbind(glm_perf_90,glm_perf_90_tmp)
  }
}
glm_perf_90$Days_before<-"90"
knitr::kable(head(glm_perf_90 %>% select(Model,Days_before,folds,times,AUC) %>% unique()))
```

### Model tuning and evaluation - Naive bayes
```{r warning=F,message=F}
nb_perf_90<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    nb_perf_90_tmp<-nb_tune_eval(training_90,test_90,i,k,seed)
    nb_perf_90<-rbind(nb_perf_90,nb_perf_90_tmp)
  }
}
nb_perf_90$Days_before<-"90"
knitr::kable(head(nb_perf_90 %>% select(Model,Days_before,folds,times,AUC) %>% unique()))
```

### Model tuning and evaluation - random forest
```{r warning=F,message=F}
rf_perf_90<-NULL
rf_tree_90<-NULL
rf_imp_90<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    rf_temp_com<-rf_tune_eval(training_90,test_90,i,k,seed)
    rf_perf_90<-rbind(rf_perf_90,rf_temp_com[[1]])
    rf_tree_90<-rbind(rf_tree_90,rf_temp_com[[2]])
    rf_imp_90<-rbind(rf_imp_90,rf_temp_com[[3]])
  }
}
rf_perf_90$Days_before<-"90"
knitr::kable(head(rf_perf_90 %>% select(Model,Days_before,folds,times,AUC) %>% unique()))
```

## Results
### Prediction model evaluation
```{r}
AUC <- rbind(glm_perf_90,nb_perf_90,rf_perf_90) %>% dplyr::select(Model,Days_before,folds,times,AUC) %>% unique()
AUCDF<-AUC %>% group_by(Model,Days_before) %>% 
  summarise(Count=n(),Mean=round(mean(AUC),digit=3),
            SE=round(sd(AUC)/150,digit=4))
knitr::kable(AUCDF)

summary(aov(AUC~Model,data=AUC[Days_before=="90"]))
TukeyHSD(aov(AUC~Model,data=AUC[Days_before=="90"]))

rf_sen_75 <- rf_perf_90[sen>=0.70 & spe>0][order(sen)][, .SD[1], by=.(folds,times)]
rf_sen_75 %>% summarize(model="Rf",sen = round(mean(sen),3), spe=round(mean(spe),3), 
                        ppv=round(mean(ppv),3), npv=round(mean(npv),3))
```

### Important features for breast cancer metastasis preduction

#### Mean Decrease Gini
```{r warning=F,message=F}
knitr::kable(rf_imp_90[,.(MeanDecreaseGini=mean(MeanDecreaseGini)),by=(rn)][order(-MeanDecreaseGini)])
```

#### Number of times being the split variable   
```{r warning=F,message=F}
knitr::kable(rf_tree_90[,.N,by=`split var`][order(-N)])
```

### The effect of time in metastasis prediction 
```{r eval=F}


summary(aov(AUC~Model,data=AUC[Days_before=="60"]))
TukeyHSD(aov(AUC~Model,data=AUC[Days_before=="60"]))

summary(aov(AUC~Model,data=AUC[Days_before=="30"]))
TukeyHSD(aov(AUC~Model,data=AUC[Days_before=="30"]))

summary(aov(AUC~Days_before,data=AUC[Model=="RF"]))
TukeyHSD(aov(AUC~Days_before,data=AUC[Model=="RF"]))
```