---
title: "Predicting Breast Cancer Metastasis Using Clinicopathological Data and Machine Learning Technologies"
author: Yi-Ju Tseng, Chuang-En Huang, Po-Yin Lai, Yu-Chen Sun, Chiao-Ni Wen, Hsin-Yao Wang, and Jang-Jih Lu
institute: "Chang Gung University"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```
## Set environment
### Load all librarys
```{r library.load, warning=F,message=F}
library(data.table)
library(ggplot2)
library(lubridate)
library(dplyr)
library(tidyr)
library(ROCR)
library(caret)
library(tableone)
```
### Load all functions
The ModelFunction.R can be download from [GitHub](https://github.com/DHLab-CGU/Breast-Cancer-Metastasis-Prediction/blob/master/ModelFunction.R).
```{r fun.load,warning=F,message=F}
source('ModelFunction.R')
```

## Data load and pre-process
### Load breast cancer data for model development
The clinical data is available from the Ethics Committee of the Chang Gung Memorial Hospital for researchers who meet the criteria for access to confidential data.
```{r data.load, warning=F}
TimeIndepData<-readRDS('TimeIndepData.rds')
str(dplyr::select(TimeIndepData,-ID))
TimeVariedData<-readRDS('TimeVariedData.rds')
str(dplyr::select(TimeVariedData,-ID))
```

### T stage code conversion

```{r code.conversion,warning=F}
TimeIndepData[pTi %in% c('4a','4b','4c','4d'),pTi:='8']
TimeIndepData[pTi=='3',pTi:='7']
TimeIndepData[pTi=='2',pTi:='6']
TimeIndepData[pTi=='1c',pTi:='5']
TimeIndepData[pTi=='1b',pTi:='4']
TimeIndepData[pTi=='1a',pTi:='3']
TimeIndepData[pTi=='1micro',pTi:='2']
TimeIndepData[pTi=='is',pTi:='1']
TimeIndepData[pTi=='x',pTi:=NA]
```

### Process categorical variables
```{r factor}
cols <- c("pTi", "pNi", "pMi", "Tissue.ER", "Tissue.PR", "Tissue.HER2")
TimeIndepData[, c(cols) := lapply(.SD, as.factor), .SDcols= cols]
```

## Patient characteristics
```{r tableone,warning=F}
LastData <- TimeVariedData[order(ID,LabName,-RCVDAT)][,.SD[c(1)],by=.(ID,LabName)]
LabWide <- spread(LastData[,.(ID,LabName,LabValue)],key=LabName,value=LabValue)
IndexDateGap<-LastData[,.(ID,Index_duration)][order(ID,Index_duration)][,.SD[c(1)],by=.(ID)]

LastRecordAfterOP <- inner_join(TimeIndepData,LabWide, by = "ID")%>%
  inner_join(., IndexDateGap, by='ID') 

vars <- c("OPAge","pTi","pNi","pMi","Tissue.ER","Tissue.PR","Tissue.HER2","CA153","CEA","HER2","Index_duration") 
cols <- c("pTi", "pNi", "pMi", "Tissue.ER", "Tissue.PR", "Tissue.HER2")
ALL_tableOne <- CreateTableOne(vars = vars, strata = c("IsRe"),factorVars = cols, data = LastRecordAfterOP) 
```
```{r tableone1,warning=F,message=F,echo=T,results='hide'}
ALL_tableOneDF<-data.table(as.matrix(print(ALL_tableOne,nonnormal= "Index_duration")),keep.rownames = T)
```
```{r tableone2,warning=F,message=F}
knitr::kable(ALL_tableOneDF)
```

## Build the model to predict breast cancer metastasis at least 3 months in advance

### Set parameters
```{r setpar}
seed<-3231
n_times<-50
n_folds<-3
trc<-trainControl(method = "cv", number = n_folds, 
                  classProbs=TRUE, summaryFunction = twoClassSummary)
```

### Get completed data 3 months before the index date
```{r data3months,warning=F,message=F}
LastRecordAfterOP_90_rmna_IsRe<-getDataForModel(TimeVariedData,TimeIndepData,90,"Recurrence")
LastRecordAfterOP_90_rmna_NotRe<-getDataForModel(TimeVariedData,TimeIndepData,90,"Non.Recurrence")
nrow(LastRecordAfterOP_90_rmna_IsRe)
nrow(LastRecordAfterOP_90_rmna_NotRe)
```

### 3 evaluation folds
```{r sampling,warning=F,message=F}
datalist<-generate3folds(LastRecordAfterOP_90_rmna_IsRe,LastRecordAfterOP_90_rmna_NotRe,seed)
training_90<-datalist[[1]]
test_90<-datalist[[2]]
```

### Model tuning and evaluation - logistic regression
```{r glm90,warning=F,message=F,cache=TRUE}
glm_perf_90<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    glm_perf_90_tmp<-glm_tune_eval(training_90,test_90,i,k,seed,trc)
    glm_perf_90<-rbind(glm_perf_90,glm_perf_90_tmp)
  }
}
glm_perf_90$Days_before<-"90"
```

### Model tuning and evaluation - Naive bayes
```{r nb90,warning=F,message=F,cache=TRUE}
nb_perf_90<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    nb_perf_90_tmp<-nb_tune_eval(training_90,test_90,i,k,seed,trc)
    nb_perf_90<-rbind(nb_perf_90,nb_perf_90_tmp)
  }
}
nb_perf_90$Days_before<-"90"
```

### Model tuning and evaluation - random forest
```{r rf90,warning=F,message=F,cache=TRUE}
rf_perf_90<-NULL
rf_tree_90<-NULL
rf_imp_90<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    rf_temp_com<-rf_tune_eval(training_90,test_90,i,k,seed,trc)
    rf_perf_90<-rbind(rf_perf_90,rf_temp_com[[1]])
    rf_tree_90<-rbind(rf_tree_90,rf_temp_com[[2]])
    rf_imp_90<-rbind(rf_imp_90,rf_temp_com[[3]])
  }
}
rf_perf_90$Days_before<-"90"
```

## Results
### Prediction model evaluation
```{r AUC90}
AUC <- rbind(glm_perf_90,nb_perf_90,rf_perf_90) %>% dplyr::select(Model,Days_before,folds,times,AUC) %>% unique()
AUCDF<-AUC %>% group_by(Model,Days_before) %>% 
  summarise(Count=n(),Mean=round(mean(AUC),digit=3),
            SE=round(sd(AUC)/n(),digit=4))
knitr::kable(AUCDF)

summary(aov(AUC~Model,data=AUC[Days_before=="90"]))
TukeyHSD(aov(AUC~Model,data=AUC[Days_before=="90"]))

rf_sen_75 <- rf_perf_90[sen>=0.70 & spe>0][order(sen)][, .SD[1], by=.(folds,times)]
rf_sen_75 %>% summarize(model="Rf",sen = round(mean(sen),3), spe=round(mean(spe),3), 
                        ppv=round(mean(ppv),3), npv=round(mean(npv),3))
```

### Important features for breast cancer metastasis preduction

#### Mean Decrease Gini
```{r mdg,warning=F,message=F}
knitr::kable(
  rf_imp_90[,.(MeanDecreaseGini=mean(MeanDecreaseGini)),by=(rn)][order(-MeanDecreaseGini)] %>% 
    head(10), 
  row.names=T)
```

#### Number of times being the split variable   
```{r split,warning=F,message=F}
knitr::kable(rf_tree_90[,.N,by=`split var`][order(-N)] %>% head(10),row.names=T)
```

### The effect of time in metastasis prediction 
#### Build and evaluate 60 days model
```{r model60,warning=F,message=F,cache=TRUE}
LastRecordAfterOP_60_rmna_IsRe<-getDataForModel(TimeVariedData,TimeIndepData,60,"Recurrence")
LastRecordAfterOP_60_rmna_NotRe<-getDataForModel(TimeVariedData,TimeIndepData,60,"Non.Recurrence")

datalist60<-generate3folds(LastRecordAfterOP_60_rmna_IsRe,LastRecordAfterOP_60_rmna_NotRe,seed)
training_60<-datalist60[[1]]
test_60<-datalist60[[2]]


glm_perf_60<-NULL
nb_perf_60<-NULL
rf_perf_60<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    glm_perf_60_tmp<-glm_tune_eval(training_60,test_60,i,k,seed,trc)
    glm_perf_60<-rbind(glm_perf_60,glm_perf_60_tmp)
  }
}
glm_perf_60$Days_before<-"60"

for (k in 1:n_times){
  for (i in 1:n_folds){
    nb_perf_60_tmp<-nb_tune_eval(training_60,test_60,i,k,seed,trc)
    nb_perf_60<-rbind(nb_perf_60,nb_perf_60_tmp)
  }
}
nb_perf_60$Days_before<-"60"

for (k in 1:n_times){
  for (i in 1:n_folds){
    rf_temp_com<-rf_tune_eval(training_60,test_60,i,k,seed,trc)
    rf_perf_60<-rbind(rf_perf_60,rf_temp_com[[1]])
  }
}
rf_perf_60$Days_before<-"60"
```
#### Build and evaluate 30 days model
```{r model30,warning=F,message=F,cache=TRUE}
LastRecordAfterOP_30_rmna_IsRe<-getDataForModel(TimeVariedData,TimeIndepData,30,"Recurrence")
LastRecordAfterOP_30_rmna_NotRe<-getDataForModel(TimeVariedData,TimeIndepData,30,"Non.Recurrence")

datalist30<-generate3folds(LastRecordAfterOP_30_rmna_IsRe,LastRecordAfterOP_30_rmna_NotRe,seed)
training_30<-datalist30[[1]]
test_30<-datalist30[[2]]

glm_perf_30<-NULL
nb_perf_30<-NULL
rf_perf_30<-NULL
for (k in 1:n_times){
  for (i in 1:n_folds){
    glm_perf_30_tmp<-glm_tune_eval(training_30,test_30,i,k,seed,trc)
    glm_perf_30<-rbind(glm_perf_30,glm_perf_30_tmp)
  }
}
glm_perf_30$Days_before<-"30"

for (k in 1:n_times){
  for (i in 1:n_folds){
    nb_perf_30_tmp<-nb_tune_eval(training_30,test_30,i,k,seed,trc)
    nb_perf_30<-rbind(nb_perf_30,nb_perf_30_tmp)
  }
}
nb_perf_30$Days_before<-"30"

for (k in 1:n_times){
  for (i in 1:n_folds){
    rf_temp_com<-rf_tune_eval(training_30,test_30,i,k,seed,trc)
    rf_perf_30<-rbind(rf_perf_30,rf_temp_com[[1]])
  }
}
rf_perf_30$Days_before<-"30"
```
#### Compare models
```{r time.model,warning=F,message=F}
AUCTime <- rbind(glm_perf_90,nb_perf_90,rf_perf_90,
                 glm_perf_60,nb_perf_60,rf_perf_60,
                 glm_perf_30,nb_perf_30,rf_perf_30) %>% 
  dplyr::select(Model,Days_before,folds,times,AUC) %>% unique()
AUCDFTime<-AUCTime %>% group_by(Model,Days_before) %>% 
  summarise(Count=n(),Mean=round(mean(AUC),digit=3),
            SE=round(sd(AUC)/n(),digit=4))
knitr::kable(AUCDFTime)
aov60<-summary(aov(AUC~Model,data=AUCTime[Days_before=="60"]))
knitr::kable(aov60[[1]])
Tukey60<-TukeyHSD(aov(AUC~Model,data=AUCTime[Days_before=="60"]))
knitr::kable(Tukey60$Model)
aov30<-summary(aov(AUC~Model,data=AUCTime[Days_before=="30"]))
knitr::kable(aov30[[1]])
Tukey30<-TukeyHSD(aov(AUC~Model,data=AUCTime[Days_before=="30"]))
knitr::kable(Tukey30$Model)

aovRF<-summary(aov(AUC~Days_before,data=AUCTime[Model=="RF"]))
knitr::kable(aovRF[[1]])
TukeyRF<-TukeyHSD(aov(AUC~Days_before,data=AUCTime[Model=="RF"]))
knitr::kable(TukeyRF$Model)

```

### Contact

Please feel free to contact `yjtseng [at] mail.cgu.edu.tw` or **open an issue on GitHub** if you have any questions.